%\documentclass[12pt,a4paper,english]{article}%
\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[11pt,a4paper,english]{amsart}
\input{/Users/jcmunozmora/Documents/Papers/syntaxis/head.tex}
\inputpaths{/Users/jcmunozmora/Documents/data/Tobon-Munoz-Dagust/tables/}
\graphicspath{{maps/}{/Users/jcmunozmora/Documents/data/Tobon-Munoz-Dagust/figures/}}
\usepackage{catchfilebetweentags}
\begin{document}


\title[The institutional dimension of illicit behavior]{The institutional dimension of illicit behavior: \\Formal land property rights and illegal crops in Colombia.}\singlespacing    
\author[Tob\' on \& Mu\~noz-Mora \& d'Anjue (2013)]{Santiago Tob\'on-Zapata \and Juan Carlos Mu\~noz-Mora \and Jess d'Anjue} 
\thanks{We gratefully acknowledge funding from the European Union Seventh Framework Programme (FP7/2007-2013) under grant agreement no: 263905 (TAMNEAC). \\
Corresponding author’s contact information: Juan Carlos Mu\~noz-Mora. European Center for Advanced Research in Economics and Statistics -ECARES . Universit\'e Libre de Bruxelles. 50, Avenue Roosevelt CP 139, 1050 Brussels, Belgium  e-mail:\textit{juan.carlos.munoz.mora@ulb.ac.be} \\
Co-authors contact information: Santiago Tob\'on-Zapata, Universit\'e Catholique de Louvain. Department of Economics, Coll\'ege L. H. Dupriez, 3 Place Montesquieu B- 1348 Louvain-la-Neuve. Email: \textit{santiagotobon@gmail.com}.  Jesse Willem d'Anjou.Center for European Research in Microfinance, Universit\'e Libre de Bruxelles. 50, Avenue Roosevelt CP 114, 1050 Brussels, Belgium, e-mail:\textit{jesse.w.d.anjou@gmail.com} }

\date{\today - \emph{Draft Version - Comments Welcome}}
\maketitle
\begin{abstract} \singlespacing    
This propose explores the different mechanism of rent-seeking thou, using a unique panel data set for Colombian municipalities. \\ 
 \textbf{Keywords:} 
\end{abstract}

\section{Empirical framework}



\subsection{Identification Strategy}

\subsection{Descriptive Statistics}

\section{Empirical results}

\subsection{Baseline results} 


\subsection{Robustness analysis} 


\section{Conclusions}

%------------------ REFERENCES -------------- %



%------------------ APENDIX -------------- %
\newpage
\newgeometry{left=1in,right=1in,top=1in,bottom=1in} 
\inputpaths{/Users/jcmunozmora/Documents/Papers/Tobon-Munoz-Danjou/Version_04-13/}
\ExecuteMetaData[tables-figures.tex]{mytag}

\end{document}