\begin{enumerate} \footnotesize
\item Abadie, A. and Imbens, G., 2006. ``Sample Properties of Matching Estimators for Average Treatment Effects,'' Econometrica 74(1), pp. 235-267.


\item Andersen, T.G. \& S\o rensen, B.E., 1996. ``GMM estimation of a stochastic volatility model: a Monte Carlo study,'' Journal of Business and Economic Statistics 14, pp. 328-352.



\item Arellano, M. \& Honore, B., 2001. ``Panel data models: some recent developments,'' In: Heckman, J.J., Leamer, E.E. (editors) Handbook of Econometrics. Elsevier, Vol. 5, chapter 53, pp. 3229-3296.

\item Baltagi, B.H., 2008. Econometric Analysis of Panel Data. Fourth Edition, Chichester. John Wiley \& Sons.

\item Beck, T. \& Levine, R., 2004. ``Stock markets, banks, and growth: panel evidence,'' Journal of Banking
and Finance 28, pp. 423-442.



\item Bond, S., 2002. ``Dynamic panel data models:
a guide to micro data methods and practice,'' Portuguese Economic Journal 1, pp. 141-162.

\item Bowsher, C.G., 2002. ``On testing overidentifying restrictions in dynamic panel data models,'' Economics Letters 77, pp. 211-220.


\item Christiaensen, L., Demery, L. \& Kuhl, J., 2011. ``The (evolving) role of agriculture in poverty reduction: An empirical perspective,'' Journal of Development Economics 96, pp. 239-254.


\item Dammert, A.C., 2008. ``Child Labor and Schooling Response to Changes in Coca Production in Rural Peru,'' Journal of Development Economics 86, pp. 164-180.

.

\item D\'{i}az, A.M. \& S\'{a}nchez, F., 2004. ``A geography of illicit crops and armed conflict in Colombia,'' Working Paper CEDE 002766, University of the Andes, Bogot\'{a}, pp. 1-78.

\item Hansen, L.P., 1982. ``Large sample properties of generalized method of moments estimators,'' Econometrica 50, 1029-1054.

\item Heid, B., Langer, J. \& Larch, M., 2012. ``Income and democracy: Evidence from system GMM estimates,'' Economics Letters 116, pp. 166-169.

\item Holtz-Eakin, D., Newey, W. \& Rosen, H.S., 1989.
``Implementing Causality Tests with Panel Data, with an Example from Local Public Finance,'' NBER Technical Working Papers 0048, National Bureau of Economic Research, Inc.


\item Newey, W.K., 1985. ``Maximum Likelihood Specification Testing and Conditional Moment Tests,'' Econometrica 53, pp. 1047-1070.



\item Roodman, D., 2009a. ``A note on the theme of too many instruments,'' Oxford Bulletin of Economics and Statistics 71, pp. 135-158.

 \item Roodman, D., 2009b. ``How to do xtabond2: an introduction to difference and system GMM in Stata,'' Stata Journal 9, pp. 86-136.



\item Windmeijer, F., 2005. ``A finite sample correction for the variance of linear efficient two-step GMM estimators,'' Journal of Econometrics 126, pp. 25-51.

\item  Ziliak, J.P., 1997. ``Efficient estimation with panel data when instruments are predetermined: an empirical comparison of moment-condition estimators,'' Journal of Business and Economic Statistics 16, pp. 419-431.

\end{enumerate}